---
title: "Data Preparation"
author: "Tyler Pelletier"
date: "1/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
In this script we are going to read in the data from the summer Pathogens-Pachyderms expedition in CSV format and combine it into a single data set. Then we will figure out how to do some checks on the data for consistency and accuracy. (AKA: we'll do some research and ask John)

First we have to read in the data that was digitized by each person. Then I'm checking to see what it looks like. The head and tail functions tell me how many lines I need to skip at the top of the data and whether I have a tail of empty rows in each data set. I can also check for any errors in importing the data.

I noticed that I didn't like some of the column names for certain rows. So I changed them either to remove units or to make them more intuitive.

I then used the str function to check the attributes of the data. I may need to set the attributes of some of the columns, such as the date column.
```{r Lucas_Data}
setwd("/Users/tylerpelletier/Desktop/pathogens-pachyderms/Path-Pachy CSV/")

LM <- read.csv("Lucas_Summer_2021.csv", header = T)
head(LM)
tail(LM) #no NA tail

my_names <- c("Expedition", "Entry", "Entry_Date", "File_Name", "Collection_Date", "Note_Taker", "Project", "Plot", "Grid", "Plant_tag", "X_coord", "Y_coord", "Tool", "POM", "DBH", "Height", "Type", "Note")
names(LM)[1:18] <- my_names[1:18]

str(LM)

#Here I change these columns from factors to dates
LM$Entry_Date <- as.Date(LM$Entry_Date, format = "%d/%m/%y") 
LM$Collection_Date <- as.Date(LM$Collection_Date, format = "%d/%m/%y")
str(LM)
```

```{r Tyler_data}
TP <- read.csv("Tyler_Summer_2021.csv", header = T, skip = 1)
names(TP)[1:18] <- my_names[1:18]

head(TP)
tail(TP) #no NA tail

str(TP)

TP$Entry_Date <- as.Date(TP$Entry_Date, format = "%d/%m/%y")
TP$Collection_Date <- as.Date(TP$Collection_Date, format = "%d/%m/%y")
str(TP)
```

With Jessica's data, we see that there is a large tail of blanks/NA's coming off the end of her data set so we have to remove it. We can't use na.omit for this because it will remove entire rows if there is one blank column in them, such as the note section. Instead, I found where the NA tail started and ended and deleted those specific rows.
```{r Jessica_first_data}
Jessica.1 <- read.csv("Jessica.1_Summer_2021.csv", header = T, skip = 2)
names(Jessica.1)[1:18] <- my_names[1:18]

head(Jessica.1)
tail(Jessica.1) #NA tail present at line 915

Jessica.1 <- Jessica.1[,-19] #removing extra column
head(Jessica.1) 
J1 <- Jessica.1[-c(915:1288), ]
tail(J1) #tail is removed

str(J1)

J1$Entry_Date <- as.Date(J1$Entry_Date, format = "%d/%m/%y")
J1$Collection_Date <- as.Date(J1$Collection_Date, format = "%d/%m/%y")
str(J1)
```

```{r Jessica_second_data}
Jessica.2 <- read.csv("Jessica.2_Summer_2021.csv", header = T, skip = 2)
names(Jessica.2)[1:18] <- my_names[1:18]

head(Jessica.2)
tail(Jessica.2) #NA tail present at line 519

Jessica.2 <- Jessica.2[,-19] #removing extra column
head(Jessica.2) 

J2 <- Jessica.2[-c(519:526), ]
tail(J2) #tail is removed

str(J2)

J2$Entry_Date <- as.Date(J2$Entry_Date, format = "%d/%m/%y")
J2$Collection_Date <- as.Date(J2$Collection_Date, format = "%d/%m/%y")
str(J2)
```

Aubrey's data has an extra column in it, so I removed it.
```{r Aubrey_data}
Aubrey <- read.csv("Aubrey_Summer_2021.csv", header = T, skip = 2)
names(Aubrey)[1:18] <- my_names[1:18]

head(Aubrey) #extra column present
tail(Aubrey) #no NA tail

AK <- Aubrey[,-19] #removing extra column
head(AK) 

str(AK)

AK$Entry_Date <- as.Date(AK$Entry_Date, format = "%d/%m/%y")
AK$Collection_Date <- as.Date(AK$Collection_Date, format = "%d/%m/%y")
str(AK)
```

```{r Sara_data}
Sara <- read.csv("Sara_Summer_2021.csv", header = T)
names(Sara)[1:18] <- my_names[1:18]

head(Sara)
tail(Sara) #NA tail present at line 1051

SH <- Sara[-c(1051:1225), ]
tail(SH) #tail is removed

str(SH)

SH$Entry_Date <- as.Date(SH$Entry_Date, format = "%d/%m/%y")
SH$Collection_Date <- as.Date(SH$Collection_Date, format = "%d/%m/%y")
str(SH)
```

```{r Enikoe_data}
Eni <- read.csv("Enikoe_Summer_2021.csv", header = T, skip = 2)
names(Eni)[1:18] <- my_names[1:18]

head(Eni)
tail(Eni) #NA tail present at line 560

Eni <- Eni[,-19] #removing extra column
head(Eni) 

EB <- Eni[-c(560:580), ]
tail(EB) #tail is removed

str(EB)

EB$Entry_Date <- as.Date(EB$Entry_Date, format = "%d/%m/%y")
EB$Collection_Date <- as.Date(EB$Collection_Date, format = "%d/%m/%y")
str(EB)
```

```{r Mishka_data}
Mishka <- read.csv("Mishka_Summer_2021.csv", header = T, skip = 2)
names(Mishka)[1:18] <- my_names[1:18]

head(Mishka) #extra column present
tail(Mishka) #no NA tail present (:

HM <- Mishka[,-19] #removing extra column
head(HM) 

str(HM)

HM$Entry_Date <- as.Date(HM$Entry_Date, format = "%d/%m/%y")
HM$Collection_Date <- as.Date(HM$Collection_Date, format = "%d/%m/%y")
str(HM)
```

Here I am combining the data sets vertically by columns with the same name.
```{r Combining_Data}
m1 = rbind(TP, LM)
m2 = rbind(m1, J1)
m3 = rbind(m2, J2)
m4 = rbind(m3, AK)
m5 = rbind(m4, SH)
m6 = rbind(m5, EB)
Full = rbind(m6, HM)
```

```{r}
str(Full)
Full$Expedition = "Summer_2021" 

levels(Full$Project)
levels(Full$Type)
levels(Full$Type)


```




